---
output: github_document
references:
- id: dewreede
  title: mstate: An R Package for the Analysis of Competing Risks and Multi-State Models
  author:
  - family: de Wreede
    given: Liesbeth C
  - family: Fiocco
    given: Marta
  - family: Putter
    given: Hein
  container-title: Journal of Statistical Software
  volume: 38
  number: i07
  type: article-journal
  issued:
    year: 2011
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# rdes

A fast implementation of discrete event simulation for R with parametric transition hazards. It is designed to be used with semi-Markov multi-state models of healthcare data, but can be used for any system that can be discretised into the arrival into states at certain time-points, with parametric families providing appropriate fits for these transition rates. It currently only has the ability for *estimating transition probabilities*, but additional features are in the pipeline, including:

  - Estimate length of stay
  - Provide full cohort simulation

Similar features are already provided by the excellent `flexsurv` R package, however, there are some limitations with their implementation, not least they it is rather slow to run. Further issues are described below in *Examples*.

## Installation

You can install rdes from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("stulacy/RDES")
```

### Windows

Note that since the simulation engine is written in C++, Windows users will need to have `Rtools` installed, which can be obtained from [https://cran.r-project.org/bin/windows/Rtools/](https://cran.r-project.org/bin/windows/Rtools/), and then set to the right path with:

```{r, eval=F}
devtools::find_rtools()
```

Once `Rtools` is setup run `devtools::install_github("stulacy/RDES")`.

## Examples

This section will demonstrate how to use `rdes` for estimating transition probabilities from a multi-state model, and the additional features that it provides over using `flexsurv::pmatrix.simfs`.

### Setup

 This guide assumes familiarity with multi-state modelling in R, and this section in particular glosses over the details and just prepares models and data in order to demonstrate the features of `rdes`. If you are unfamiliar with multi-state modelling then I would recommend reading [@dewreede] or the [tutorial by Putter](https://cran.r-project.org/web/packages/mstate/vignettes/Tutorial.pdf), both of whom are invovled with the development of the `mstate` package.
 
For these examples the `ebmt3` data set from `mstate` will be used. This provides a simple illness-death model of patients following transplant. The initial state is patient having received transplantation, *pr* referring to platelet recovery (the 'illness'), with relapse-free-survival (*rfs*) being the only sink state.

```{r}
library(mstate)
data(ebmt3)
head(ebmt3)
```

`mstate` provides a host of utility functions for working with multi-state models. For example, the `trans.illdeath()` function provides the required transition matrix for this state structure (`transMat` should be used when greater flexibility is required).

```{r}
tmat <- trans.illdeath()
tmat
```

The final data preparation step is to form the data from a wide format (each row corresponding to a patient) to a long format, where each row represents a potential patient-transition. The `msprep` function from `mstate` handles this for us. We'll keep both the `age` and `dissub` covariates in this example.

```{r}
long <- msprep(c(NA, 'prtime', 'rfstime'), c(NA, 'prstat', 'rfsstat'), ebmt3, tmat, keep=c('age', 'dissub'))
head(long)
```

Clock-reset Weibull models will be fitted to these 3 transitions, which are semi-Markov models. Simulation is therefore needed to obtain transition probabilities as the Kolmogorov forward differential equation is no longer valid with the violation of the Markov assumption. We are going to assume that the baseline hazard isn't proportional between transitions and there are no shared transition effects for simplicity's sake.

```{r}
library(flexsurv)
models <- lapply(1:3, function(i) flexsurvreg(Surv(time, status) ~ age + dissub, data=long, dist='weibull'))
```

### Estimating transition probabilities

We'll estimate the transition probabilities of an individual with the covariates `age=20-40` and `dissub=AML` at 1 year after transplant.

```{r}
newdata <- data.frame(age=factor("20-40", levels=c("<=20", "20-40", ">40")),
                      dissub=factor("AML", levels=c("AML", "ALL", "CML")))
```

```{r example1}
library(rdes)
predict_transitions(models, newdata, tmat, times=365)
```

Note that this gives very similar responses to `pmatrix.simfs`.

```{r example2}
pmatrix.simfs(models, tmat, newdata=newdata, t=365)
```

Confidence intervals can be constructed in the same fashion as `pmatrix.simfs`, using draws from the multivariate-normal distribution of the parameter estimates.

```{r example3}
predict_transitions(models, newdata, tmat, times=365, ci=TRUE, M=10)
```

Which gives rather different results to those obtained from `pmatrix.simfs` which seem to be too wide and the estimate value is far different to that obtained when run without CIs.

```{r example4}
pmatrix.simfs(models, tmat, newdata=newdata, t=365, ci=TRUE, M=9)
```

Note that on a single individual the speed-up isn't present, with `rdes` taking 4 times longer than `flexsurv`, although the difference between 1.2s and 0.4s isn't that noticeable in interactive work. The main benefit of `rdes` comes when estimating more involved probabilities, as will be demonstrated next.

```{r}
library(microbenchmark)
microbenchmark("rdes"=predict_transitions(models, newdata, tmat, times=365),
               "flexsurv"=pmatrix.simfs(models, tmat, newdata=newdata, t=365), times=10)
```

### Estimating probabilities at multiple times

Frequently, it is useful to be able to estimate transition probabilities at multiple time-points, in order to build up a picture of an individual's disease progression for plotting. `pmatrix.simfs` only allows a single value to be passed in, and so it is up to the user to manually iterate through the time-scale. In the example below we will calculate transition probabilities at yearly intervals for 9 years.

```{r example5}
predict_transitions(models, newdata, tmat, times=seq(9)*365)
```

In `pmatrix.simfs` it is up to the user to add the times to the output to make it interpretable. Again, the probabilities agree with each other.

```{r example2}
do.call('rbind', lapply(seq(9)*365, function(t) {
    pmatrix.simfs(models, tmat, newdata=newdata, t=t)
}))
```

By removing this boilerplate code, the speed increase of `rdes` starts to show, with calculating 8 additional time-points only increasing the runtime from 1.2s to 1.6, while `flexsurv` has gone from 0.3s to 2.7s.

```{r benchmarkmultipletimes}
microbenchmark("rdes"=predict_transitions(models, newdata, tmat, times=seq(9)*365),
               "flexsurv"={do.call('rbind', lapply(seq(9)*365, function(t) {
                            pmatrix.simfs(models, tmat, newdata=newdata, t=t)}))
               }, times=10)
```

### Changing start time

`pmatrix.simfs` limits the user to specifying a start time of $s=0$. In `rdes` this is fully customisable. For example, the call below shows estimates the 1-year transition probabilities conditioned on the individual being alive at 6 months (technically it also calculates the transition probabilities conditioned on being dead at 6 months in the third row, but these aren't helpful). Notice how the probabilities of being dead at 1 year have decreased as a result.

```{r example5}
predict_transitions(models, newdata, tmat, times=365, start_times = 365)
```

Multiple start times can be provided too, such as the quarterly predictions below.

```{r example5}
predict_transitions(models, newdata, tmat, times=365, start_times = c(365 * 0.25, 365 * 0.5, 365 * 0.75))
```

Finally, any combination of number of start times and end times can be specified **provided that all values in `start_times` are less than $min(\text{times})$**. 

```{r example5}
predict_transitions(models, newdata, tmat, times=seq(2)*365, start_times = c(365 * 0.25, 365 * 0.5, 365 * 0.75))
```

Note that obtaining these additional probabilities does not increase the runtime of the function.

```{r}
microbenchmark("time"=predict_transitions(models, newdata, tmat, times=seq(2)*365, start_times = c(365 * 0.25, 365 * 0.5, 365 * 0.75)),
               times=10)
```

### Multiple individuals

Estimating transition probabilities for multiple individuals is a common need, to see how the outcomes differ for patients with different characteristics. `rdes` simply handles multiple rows supplied to `newdata`.

```{r}
newdata_multi <- data.frame(age=factor(c("20-40", ">40"), levels=c("<=20", "20-40", ">40")),
                            dissub=factor(c("AML", "CML"), levels=c("AML", "ALL", "CML")))
```

```{r exampleinds1}
predict_transitions(models, newdata_multi, tmat, times=365)
```

As with multiple times, `pmatrix.simfs` only works with a single individual at a time.

```{r exampleinds2, erro=T}
pmatrix.simfs(models, tmat, newdata=newdata_multi, t=365)
```

And the user has to manually iterate through each new individual they would like to estimate transition probabilities for.

```{r exampleinds3}
do.call('rbind', lapply(seq(nrow(newdata_multi)), function(i) {
    pmatrix.simfs(models, tmat, newdata=newdata_multi[i, ], t=365)
}))
```

### Time-dependent covariates

The Markov assumption has already been violated by the use of a clock-reset time-scale which is why we are using simulation in the first place. We can therefore add an other violation without it affecting our methodology. Owing to the use of clock-reset, the model does not take time-since-transplant into account for patients who have platelet recovery. This could be an important prognostic factor in that individual's survival. Similar scenarios are common in multi-state modelling, and are termed `state-arrival` times. We'll make a new set of models, where the transition from `pr` to `rfs` (transition 3) takes time-since-transplant into account. This information is already held in the `Tstart` variable produced by `msprep`.

```{r}
models_arrival <- lapply(1:3, function(i) {
    if (i == 3) {
        flexsurvreg(Surv(time, status) ~ age + dissub + Tstart, data=long, dist='weibull')
    } else {
        
        flexsurvreg(Surv(time, status) ~ age + dissub, data=long, dist='weibull')
    }
})
```

Looking at the coefficient for this variable and it does seem to be prognostic for time-to-rfs.

```{r}
models_arrival[[3]]
```

This variable needs to be included in the data set supplied to `newdata` with its **initial value**.

```{r}
newdata_arrival <- data.frame(age=factor("20-40", levels=c("<=20", "20-40", ">40")),
                              dissub=factor("AML", levels=c("AML", "ALL", "CML")),
                              Tstart=0)
```

Simply specify which variables in `newdata` are time-dependent, that is they increment at each transition along with the current clock value. This is particuarly useful for modelling patient age at each state entry, rather than at the starting state. Notice how this slightly changes the transition probabilities from the illness state compared to the example below that omits the `tcovs` argument.

```{r exampleinds1}
predict_transitions(models_arrival, newdata_arrival, tmat, times=365, tcovs='Tstart')
```

```{r exampleinds1}
predict_transitions(models_arrival, newdata_arrival, tmat, times=365)
```

This is implemented in `pmatrix.simfs`, but `tcovs` actually have no impact on the transition probabilities, as evidenced below.

```{r tcovs1}
pmatrix.simfs(models_arrival, tmat, newdata=newdata_arrival, t=365, tcovs='Tstart')
```

```{r tcovs2}
pmatrix.simfs(models_arrival, tmat, newdata=newdata_arrival, t=365)
```

### Mixture of distributions

Sometimes flexibility in the model structure is required, such that not every transition is required to use the same distribution. This is shown below, where for each transition a different distribution family is chosen. This could be useful if some of the transitions have fewer patients and require a more simple model such as an exponential, or are restricted by the use of a particular model from literature. If prediction is the goal then it could be the case that using different distributions on each transition offers better fit.

```{r}
models_mix <- lapply(1:3, function(i) {
    if (i == 1) {
        flexsurvreg(Surv(time, status) ~ age + dissub, data=long, dist='weibull')
    } else if (i == 2) {
        flexsurvreg(Surv(time, status) ~ age + dissub, data=long, dist='exp')
    } else {
        flexsurvreg(Surv(time, status) ~ age + dissub, data=long, dist='lnorm')
    }
})
```

`rdes` handles these cases with no problems, although currently the supported distributions are:

  - Weibull
  - Gamma
  - Exponential
  - Log-normal
  - Log-logistic
  - Gompertz

```{r exampleinds1}
predict_transitions(models_mix, newdata, tmat, times=365)
```

`pmatrix.simfs` does not seem to function correctly under these situations.

```{r tcovs2}
pmatrix.simfs(models_mix, tmat, newdata=newdata, t=365)
```

## Upcoming features

Length fof stay
Cohort wide simulation

Currently, it is mostly used as the simulation engine behind a web-app written in Shiny that [provides a graphical interface for the entire multi-state modelling process.](https://github.com/stulacy/RDES-Shiny)

TODO:
  - add maths
  - Replace start times and end times with the mathematical symbols for these
  
## Bibliography